---
title: "Nucleosome Positioning"
output:
  BiocStyle::html_document:
    toc: true
bibliography: rjmcmc.bibtex
vignette: >
  %\VignetteIndexEntry{Nucleosome Positioning}
  %\VignettePackage{RJMCMCNucleosomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<br />
**Package**: `r Rpackage("RJMCMCNucleosomes")`<br />
**Authors**: `r packageDescription("RJMCMCNucleosomes")[["Author"]]`<br />
**Version**: `r packageDescription("RJMCMCNucleosomes")$Version`<br />
**Compiled date**: `r Sys.Date()`<br />
**License**: `r packageDescription("RJMCMCNucleosomes")[["License"]]`<br />

# Licensing and citing

The **RJMCMCNucleosomes** package and the underlying **RJMCMCNucleosomes** code 
are distributed under the Artistic license 2.0. You are free to use and 
redistribute this software. 

If you use this package for a publication, we would ask you to cite the
following:

>Samb R, Khadraoui K, Belleau P, et al. (2015) Using informative Multinomial-Dirichlet prior in a t-mixture with reversible jump estimation of nucleosome positions for genome-wide profiling. Statistical Applications in Genetics and Molecular Biology. Volume 14, Issue 6, Pages 517-532, ISSN (Online) 1544-6115, ISSN (Print) 2194-6302, December 2015, <a href="http://dx.doi.org/10.1515/sagmb-2014-0098">doi:10.1515/sagmb-2014-0098</a>


# Introduction

Global gene expression patterns are established and maintained by the 
concerted actions of transcription factors (TFs) and the proteins that 
constitute chromatin. The key structural element of chromatin is the 
nucleosome, which consists of an octameric histone core wrapped by 146 bps 
of DNA and connected to its neighbour by approximately 10-80 pbs of linker 
DNA [@Polishko2012].

The literature on nucleosome positioning commonly focuses on frequentist 
inferences within parametric approaches (see for instance @Chen2010 and
@Xi2010).  In those works, the detection of nucleosome positions is done 
using a hidden Markov model with an assumed known order.

The **RJMCMCNucleosomes** package is an implementation of a fully Bayesian 
hierarchical model for profiling of nucleosome positions based on 
high-throughput short-read data (MNase-Seq data). The implementation is based 
on a strategy which incorporates four aspects. First, it jointly models local 
concentrations of directional reads. Second, it uses a Multinomial-Dirichlet 
model in the construction of an informative prior distribution coupled to a 
t-mixture model with unknown degrees of freedom. Third, the number of 
nucleosomes is considered to be a random variable and refers to a prior 
distribution. Fourth, the unknown parameters are simultaneously using the 
reversible jump Markov chain Monte Carlo (RJMCMC) simulation technique 
(see for instance @Green1995 and @Richardson1997). 

Detailed information about the model can be found in the article 
mentioned in the citing section.

# Loading the RJMCMC package

As with any R package, the **RJMCMCNucleosomes** package should first be loaded 
with the following command:

```{r loadingPackage, warning=FALSE, message=FALSE} 
library(RJMCMCNucleosomes)
```


# RJMCMCNucleosomes analysis

A typical **RJMCMCNucleosomes** analysis consists of the following steps:

1. Segment the genome into candidate regions that have sufficient aligned
reads. The region cannot be larger than one chromosome.
2. Estimate nucleosome positions for each region.
3. Regroup all regions together. The final region cannot be larger than
one chromosome.
4. Post-process predictions of the regrouped region to correct certain 
predictions.


# RJMCMCNucleosomes analysis step by step

A synthetic nucleosome sample containing 100 nucleosomes (80 
well-positioned + 20 fuzzy) has been created using the 
Bioconductor package `r Biocpkg("nucleoSim")`. This synthetic sample will be 
used throughout this analysis.

```{r createSample, collapse=TRUE, message=FALSE}
## Load nucleoSim package
library(nucleoSim)

val.num       <- 50     ### Number of well-positioned nucleosomes
val.del       <- 10     ### Number of well-positioned nucleosomes to delete
val.var       <- 30     ### variance associated to well-positioned nucleosomes
val.fuz       <- 10     ### Number of fuzzy nucleosomes
val.fuz.var   <- 50     ### variance associated to fuzzy nucleosomes
val.max.cover <- 70     ### Maximum coverage for one nucleosome
val.nuc.len   <- 147    ### Distance between nucleosomes
val.len.var   <- 10     ### Variance associated to the length of the reads
val.lin.len   <- 20     ### The length of the DNA linker
val.rnd.seed  <- 100    ### Set seed when result needs to be reproducible
val.offset    <- 10000  ### The number of bases used to offset 
                        ### all nucleosomes and reads

## Create sample using a Normal distribution
sample <- nucleoSim::syntheticNucReadsFromDist(wp.num=val.num,
                                    wp.del=val.del, wp.var=val.var,
                                    fuz.num=val.del, fuz.var=val.fuz.var,
                                    max.cover=val.max.cover, 
                                    nuc.len=val.nuc.len,
                                    len.var=val.len.var, 
                                    lin.len=val.lin.len,
                                    rnd.seed=val.rnd.seed,
                                    distr="Normal", offset=val.offset)

## Create visual representation of the synthetic nucleosome sample
plot(sample)
```


## Split the analyzed region into segments

It is suggested, in order to accelerate the learning process, to split the 
analyzed region into segments to accelerate the analysis. Moreover, 
it is mandatory to analyse each chromosome separately since the 
\code{rjmcmc} function can only analyze one chromosome at the time.

Region segmentation can be done using the \code{segmentation} function. Beware
that larger is the size of a segment (parameter \code{maxLength}), the higher 
the number of iterations need to be to reach convergence during nucleosome 
predictions step.

```{r segment01, warning=FALSE, collapse=TRUE, message=FALSE} 
## Load needed packages
library(GenomicRanges)

## Transform sample dataset into GRanges object
sampleGRanges <- GRanges(seqnames = sample$dataIP$chr, 
                          ranges = IRanges(start = sample$dataIP$start, 
                                           end = sample$dataIP$end), 
                          strand = sample$dataIP$strand)

## Segment sample into candidate regions
sampleSegmented <- segmentation(dataIP = sampleGRanges, zeta = 147, 
                                delta = 40, maxLength = 1000)

## Number of segments created
length(sampleSegmented)
```


## Run RJMCMCNucleosomes for nucleosome predictions

The \code{rjmcmc} function must be run on each candidate region. As an 
example, the first candidate region is processed using a very low number 
of iterations. On real data, the number of iterations should be higher 
(easily 1000000 iterations).

```{r runRJMCMC01 , warning=FALSE, collapse=TRUE} 
## Extract the first segment 
segment01 <- sampleSegmented[[1]]

## Run RJMCMC analysis
## A higher number of iterations is recommanded for real analysis
resultSegment01 <- rjmcmc(forwardandReverseReads  = segment01,
                            nbrIterations = 100000, lambda = 3, kMax = 30,
                            minInterval = 100, maxInterval = 200, minReads = 5,
                            vSeed = 1921)

## Print the predicted nucleosomes for the first segment
resultSegment01
```


## Regroup all regions

Once all segments have been analyzed, the predicted nucleosomes can be merged 
together. Two functions are available to facilitate the merging process:

* *mergeRDSFiles* function: An array containing the name of all RDS files
to merge is passed to it
* *mergeAllRDSFilesFromDirectory* function: the name of the directory 
(relative or absolute path) containing all RDS files to merge is passed to it

Beware that segment from different chromosomes should not be merged together.

The segments of the sample created sooner have all been processed (using 
500000 iterations) and saved in RDS files. Those will now be merged together.

```{r regroup01, warning=FALSE, collapse=TRUE, message=FALSE}
## The directory containing the results of all segments
## On RDS file has been created for each segment
directory <- system.file("extdata", "demo_vignette", 
                            package = "RJMCMCNucleosomes")

## Merging predicted nucleosomes from all segments
resultsAllMerged <- mergeAllRDSFilesFromDirectory(directory)

resultsAllMerged
```


## Post-process predictions

In some cases the RJMCMC method tends to over split the distribution of reads 
for a single nucleosome. Although this characteristic increases the number 
of false positives, it is still beneficial for the region’s rich in 
nucleosomes.

A function, that merges closely positioned nucleosomes, has been implemented 
to rectify the over splitting and provide more conservative results.

The \code{postTreatment} function must be run on the entire analyzed region to
be efficient. __It should not be run on segmented results__. The function 
needs the positions of the reads used for the RJMCMC analysis.

The value of \code{extendingSize} should be kept low (bellow 20). A larger 
value could cause the possible merge of true nucleosomes. 

```{r postProcess01, collapse=TRUE, message=FALSE}
## Split reads from the initial sample data into forward and reverse subsets
reads <- GRanges(sample$dataIP)

## Number of nucleosomes before the post-treatment
resultsAllMerged$k

## Use the post-treatment function
resultsPostTreatment <- postTreatment(forwardandReverseReads = reads,
                            resultRJMCMC = resultsAllMerged,
                            extendingSize = 15,
                            chrLength = max(start(reads), end(reads)) + 1000)


## Number of nucleosomes after the post-treatment
length(resultsPostTreatment)
```

The \code{postTreatment} function can significantly reduces the number of 
nucleosomes by merging closely positioned nucleosomes.


## Visualisation of the predicted nucleosomes

Visualisation of the predicted nucleosomes, with its associated read coverage, 
is available in the **RJMCMCNucleosomes** package.

The \code{plotNucleosomes} function needs the nucleosome positions as well as 
the reads, in a \code{IRanges} format, to create a graph. When reads are used 
to predict more than one set of nucleosome positions (as examples, before and
after post-treatment or results from different software), the predictions can
be merged in a \code{list} so that all predictions can be plotted 
simultaneously.

```{r visualisation, collapse=TRUE, message=FALSE, fig.height=6.5, fig.width=8}
## Extract reads to create an IRange
reads <-IRanges(start = sample$dataIP$start, end=sample$dataIP$end)

## Merge predictions from before post-treatment and after post-treatment in 
## a list so that both results will be shown in graph
resultsBeforeAndAfter <- list(Sample = c(sample$wp$nucleopos, 
                                            sample$fuz$nucleopos),
                              BeforePostTreatment = resultsAllMerged$mu,    
                              AfterPostTreatment = resultsPostTreatment)

## Create graph using nucleosome positions and reads
## The plot will shows : 
## 1. nucleosomes from the sample
## 2. nucleosomes detected by rjmcmc() function
## 3. nucleosomes obtained after post-treament 
plotNucleosomes(resultsBeforeAndAfter, reads, 
                    names=c("Sample", "RJMCMC", "After Post-Treatment"))
```


# Session info

Here is the output of `sessionInfo()` on the system on which this document was 
compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


# References
